---
title: "Heart Rates"
author: "Valerio Gomez-Frittelli"
format: ptbl
editor: visual
---

# Data Handling

### Packages

```{r}
library(tidyverse) |> suppressMessages() |> suppressWarnings()
library(fpp3) |> suppressMessages() |> suppressWarnings()
library(forecast) |> suppressMessages() |> suppressWarnings()
library(zoo) |> suppressMessages() |> suppressWarnings()
```

## Data Loading

### Load data function

```{r}
load_data <- function(subject, biomarker) {
  paste("../Empatica_data/inputs/", subject, "_", 
        biomarker, "s", sep="") |>
    list.files(pattern = "*.csv", full.names = TRUE) |> 
    lapply(readr::read_csv) |> 
    bind_rows() |> suppressMessages() |> suppressWarnings()
}
```

### SPT04 load data call

```{r}
spt04_hr_tbl <- load_data("SPT04", "heart_rate")
```

### SPT05 load data call

```{r}
spt05_hr_tbl <- load_data("SPT05", "heart_rate")
```

## Resample to Daily Average

### Resample function

```{r}
resample_daily <- function(hr_tbl) {
  hr_tbl |> 
    mutate(day = as.Date(floor_date(timestamp_iso, "day"))) |>
    group_by(day) |>
    summarise(hr_avg = mean(pulse_rate_bpm, na.rm = TRUE))
}
```

### SPT04 resample call

```{r}
spt04_hr_daily_tbl <- resample_daily(spt04_hr_tbl)
```

### SPT05 resample call

```{r}
spt05_hr_daily_tbl <- resample_daily(spt05_hr_tbl)
```

## Interpolate Missing Data

### Interpolate function

```{r}
interpolate_missing <- function(hr_daily_tbl) {
  hr_daily_tbl_full <- hr_daily_tbl %>%
    complete(day = seq(min(day), max(day), by = "day"))
  
  hr_daily_tbl_full$hr_avg <- na.approx(hr_daily_tbl_full$hr_avg,
                                       x = hr_daily_tbl_full$day, 
                                       rule = 2, na.rm = FALSE)
  
  return(hr_daily_tbl_full)
}
```

### SPT04 interpolate call

```{r}
spt04_hr_daily_tbl <- interpolate_missing(spt04_hr_daily_tbl)
```

### SPT05 interpolate call

```{r}
spt05_hr_daily_tbl <- interpolate_missing(spt05_hr_daily_tbl)
```

## Add Phase Labellings

### SPT04 phases

```{r}
spt04_hr_daily_tbl <- spt04_hr_daily_tbl |> 
  mutate(phase = ifelse(
    day < as.Date("2026-02-02"), "Baseline", ifelse(
      day < as.Date("2026-03-16"), "Rehab", ifelse(
        day < as.Date("2026-03-23"), "Implant", ifelse(
          day < as.Date("2026-03-30"), "Implant Recovery", ifelse(
            day < as.Date("2026-04-20"), "Optimization", ifelse(
              day < as.Date("2026-05-18"), "Implant+Rehab", "Post-Rehab"
    ))))))
  )
```

### SPT05 phases

```{r}
spt05_hr_daily_tbl <- spt05_hr_daily_tbl |> 
  mutate(phase = ifelse(
    day < as.Date("2026-02-23"), "Baseline", ifelse(
      day < as.Date("2026-04-06"), "Rehab", ifelse(
        day < as.Date("2026-04-13"), "Implant", ifelse(
          day < as.Date("2026-04-20"), "Implant Recovery", ifelse(
            day < as.Date("2026-05-11"), "Optimization", ifelse(
              day < as.Date("2026-06-22"), "Implant+Rehab", "Post-Rehab"
    ))))))
  )
```

# Exploratory Plotting

## HR Longitudinal

### Longitudinal plotting function

```{r}
plot_hr_longitudinal <- function(hr_daily_tbl, subject) {
  hr_daily_tbl |>
    ggplot(aes(x=day, y=hr_avg)) + geom_line(aes(color=phase)) +
    labs(y="hr Daily Average", x="Day", 
         title=paste(subject, "Average hr by Day")) +
    theme(plot.title = element_text(hjust=0.5))
}
```

### SPT04 longitudinal plot call

```{r}
plot_hr_longitudinal(spt04_hr_daily_tbl, "SPT04")
```

### SPT05 longitudinal plot call

```{r}
plot_hr_longitudinal(spt05_hr_daily_tbl, "SPT05")
```

# Sample-size backtesting

### Helper functions

```{r}
# Fit models on current training set
fit_models <- function(train_tsibble) {
  train_tsibble |>
    model(
      'Mean' = MEAN(hr_avg),
      # Try removing trend/seasonality
      'ETS' = ETS(hr_avg ~ error("A") + trend("A") + season("A")),
      'ARIMA' = ARIMA(hr_avg ~ pdq() + PDQ(),
                      stepwise=FALSE, greedy=FALSE),
      'RW' = RW(hr_avg)
    )
}

# Evaluate trained models on current test set
evaluate_models <- function(forecasts, full_tsibble) {
  forecasts |>
    accuracy(full_tsibble) |>
    select('.model', 'MAPE') |>
    mutate(Model = .model) |>
    select(-.model) |>
    arrange(MAPE)
}

last_week_avg_forecast <- function(train_tsibble, test_tsibble){
  train_tsibble |>
    as_tibble() |> 
    filter(day >= max(day) - 6) |>
    pull(hr_avg) |>
    mean() %>% 
    { tibble(day = test_tsibble$day, fc = rep(., 7)) }
}

manual_mape_comp <- function(predictions_tbl, test_tsibble) {
  predictions_tbl |> left_join(as_tibble(test_tsibble), by = "day") |>
    summarise(MAPE = mean((abs(actual_hr_avg - fc)/actual_hr_avg),
                          na.rm = TRUE) * 100) |> 
    pull(MAPE)
}
```

### Single-size backtesting function

```{r}
backtest_at_size <- function(sample_size_mape_tbl, full_tsibble,
                             train_size, window_shift, abs_start_date) {
  
  # Reset windows to beginning of allowed timeframe
  train_start <- abs_start_date
  test_start <- train_start + train_size
  
  # Run backtesting over all valid train/test windows
  while(test_start + 6 <= max(full_tsibble$day)) {
    # 14-day training window
    train_tsibble <- full_tsibble |> 
      filter(day >= as.Date(train_start) & day < as.Date(test_start))
    
    # 7-day testing window
    test_tsibble <- full_tsibble |> 
      filter(day >= test_start & day < (test_start + 7)) |>
      mutate(actual_hr_avg = hr_avg) |> select(-hr_avg)
    
    # Fit new models on current train set and forecast current test week
    models <- fit_models(train_tsibble)
    forecasts <- models |> forecast(h=7)
    
    evals <- evaluate_models(forecasts, full_tsibble)
    
    ets_mape <- evals |> filter(Model == "ETS") |> pull(MAPE)
    arima_mape <- evals |> filter(Model == "ARIMA") |> pull(MAPE)
    window_avg_mape <- evals |> filter(Model == "Mean") |> pull(MAPE)
    rw_mape <- evals |> filter(Model == "RW") |> pull(MAPE)
    
    # Baseline: Last-week mean of train window
    last_week_avg_fc <- last_week_avg_forecast(train_tsibble, test_tsibble)
    last_week_avg_mape <- manual_mape_comp(last_week_avg_fc, test_tsibble)
    
    # Add row to full tibble with all mapes, sample size, and test date
    sample_size_mape_tbl <- sample_size_mape_tbl |> 
      add_row(
        # Sample size and test date jointly specify unique 'observations'
        sample_size = train_size, test_start = test_start,
        ets_mapes = ets_mape, arima_mapes = arima_mape, 
        window_avg_mapes = window_avg_mape, 
        last_week_avg_mapes = last_week_avg_mape,
        rw_mapes = rw_mape
        )
    
    # Slide train and test windows by specified amount
    train_start <- train_start + window_shift
    test_start <- test_start + window_shift
  }
  return(sample_size_mape_tbl)
}
```

### Sample-size-range backtesting function

```{r}
backtest_sample_range <- function(hr_daily_tbl, abs_start_date, 
                                  window_shift, sample_sizes) {
  #### Parameters: #####
  # abs_start_date: Earliest date to start training models
  # window_shift: How many days to shift train/test windows each iteration
  # sample_sizes: Vector of sample sizes to backtest on
  
  #### Initialization ####
  full_tsibble <- as_tsibble(hr_daily_tbl, index=day)
  
  sample_size_mape_tbl <- tibble(
    sample_size          = integer(0),
    test_start           = as.Date(character(0)),
    ets_mapes            = numeric(0),
    arima_mapes          = numeric(0),
    window_avg_mapes     = numeric(0),
    last_week_avg_mapes  = numeric(0),
    rw_mapes             = numeric(0)
  )
  
  #### Backtesting loop ####
  for (train_size in sample_sizes) {
    sample_size_mape_tbl <- backtest_at_size(
      sample_size_mape_tbl, full_tsibble, 
      train_size, window_shift, abs_start_date
      )
  }
  return(sample_size_mape_tbl)
}
```

### SPT04 sample-size-range backtesting call

```{r}
spt04_abs_start_date <- as.Date("2026-01-23")

spt04_hr_ss_bt_tbl <- backtest_sample_range(
  spt04_hr_daily_tbl, spt04_abs_start_date, 
  window_shift=7, sample_sizes=seq(14, 30)
  )
```

### SPT05 sample-size-range backtesting call

```{r}
spt05_abs_start_date <- as.Date("2026-02-18")

spt05_hr_ss_bt_tbl <- backtest_sample_range(
  spt05_hr_daily_tbl, spt05_abs_start_date, 
  window_shift=7, sample_sizes=seq(14, 30)
  )
```

# Backtesting Plotting and Evaluation

## Longitudinal Plotting

### Longitudinal helper function

```{r}
# Helper function
plot_model_longitudinal <- function(sample_size_mape_tbl, train_size, 
                                    model_col, plot_color, 
                                    subject, model_name) {
  if ((sample_size_mape_tbl |> 
      filter(sample_size==train_size) |>
      pull(model_col) |> 
      max(na.rm = TRUE)) < 100) {
    y_upper_limit <- 100
  } else {
    y_upper_limit <- NA
  }
  
  sample_size_mape_tbl |> filter(sample_size==train_size) |>
    ggplot(aes(x=test_start, y=.data[[model_col]])) + 
    geom_line(color=plot_color) +
    scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
    scale_y_continuous(limits=c(0, y_upper_limit)) +
    labs(title=paste(
      subject, model_name, 
      "MAPEs by Test Week, sample size =", 
      train_size
      ), y="MAPE", x="Testing week (start)") +
    theme(plot.title=element_text(hjust=0.5))
}
```

### Longitudinal plotting function

```{r}
plot_sample_sizes_longitudinal <- function(sample_size_mape_tbl, 
                                           sample_sizes, subject) {
  #### Parameters: ####
  # sample_sizes: Vector of sample sizes to backtest on
  # subject: A string, either "SPT04" or "SPT05", for plot titles
  model_cols <- c('ets_mapes', 'arima_mapes', 
                  'window_avg_mapes', 'last_week_avg_mapes', 'rw_mapes')
  model_names <- c("ETS", "ARIMA", "Window Mean", "Last-week Average", 
                   "Random Walk")
  plot_colors <- c("blue", "red", "green4", "yellow3", "orange")
  
  
  for (train_size in sample_sizes) {
    # Plot longitudinal mape for each model within each sample size
    for (model_idx in seq(1, 5)) {
      model_col <- model_cols[model_idx]
      model_name <- model_names[model_idx]
      plot_color <- plot_colors[model_idx]
      plot_model_longitudinal(sample_size_mape_tbl, train_size, 
                              model_col, plot_color,
                              subject, model_name) |> 
        print()
    }
  }
}
```

### SPT04 longitudinal plotting call

Suggestions for sample_size parameter:

seq(lo, hi) creates a vector with all values in the range \[lo, hi\].

unique(sample_size_mape_tbl\$sample_size) creates a vector of all backtested sample sizes.

```{r}
plot_sample_sizes_longitudinal(spt04_hr_ss_bt_tbl, 
                               sample_sizes=c(25, 26), "SPT04")
```

### SPT05 longitudinal plotting call

```{r}
plot_sample_sizes_longitudinal(spt05_hr_ss_bt_tbl, 
                               sample_sizes=c(28), "SPT05")
```

## Aggregated Plotting

### Aggregate to sample size function

```{r}
aggregate_to_sample_size <- function(sample_size_mape_tbl) {
  sample_sizes <- unique(sample_size_mape_tbl$sample_size)
  result_list <- vector("list", length(sample_sizes))
  
  for (i in seq_along(sample_sizes)) {
    ss <- sample_sizes[i]
    row <- sample_size_mape_tbl |> filter(sample_size == ss)
    
    result_list[[i]] <- tibble(
      sample_size        = ss,
      ets_agg_mape       = mean(row$ets_mapes, na.rm = TRUE),
      arima_agg_mape     = mean(row$arima_mapes, na.rm = TRUE),
      window_agg_mape    = mean(row$window_avg_mapes, na.rm = TRUE),
      last_week_agg_mape = mean(row$last_week_avg_mapes, na.rm=TRUE),
      rw_mape            = mean(row$rw_mapes, na.rm=TRUE)
    )
  }
  
  return(bind_rows(result_list))
}
```

### SPT04 aggregate to sample size call

```{r}
spt04_hr_ss_agg_tbl <- aggregate_to_sample_size(spt04_hr_ss_bt_tbl)
```

### SPT05 aggregate to sample size call

```{r}
spt05_hr_ss_agg_tbl <- aggregate_to_sample_size(spt05_hr_ss_bt_tbl)
```

### Aggregated plotting function

```{r}
plot_sample_sizes_aggregated <- function(sample_agg_tbl) {
  # Pivot to allow for plotting all five models at once with a legend
  sample_agg_tbl |>
    pivot_longer(
      cols = c(ets_agg_mape, arima_agg_mape,
               window_agg_mape, last_week_agg_mape, rw_mape),
      names_to = "model",
      values_to = "agg_mape"
    ) |>
    ggplot(aes(x = sample_size, y = agg_mape, color = model)) +
    geom_line() +
    scale_x_continuous(breaks = unique(sample_agg_tbl$sample_size)) +
    # scale_y_continuous(limits=c(0, NA)) +
    scale_color_manual(
      values = c(
        ets_agg_mape        = "blue",
        arima_agg_mape      = "red",
        window_agg_mape     = "green4",
        last_week_agg_mape  = "yellow3",
        rw_mape             = "orange"
      ),
      labels = c(
        ets_agg_mape        = "ETS",
        arima_agg_mape      = "ARIMA",
        window_agg_mape     = "Window Average",
        last_week_agg_mape  = "Last-Week Average",
        rw_mape             = "Random Walk"
      )
    ) +
    labs(x = "Sample Size", y = "MAPE", color = "Model",
         title = "Aggregated MAPE by training set size") +
    theme(plot.title=element_text(hjust=0.5))
}
```

### SPT04 aggregated plotting call

```{r}
plot_sample_sizes_aggregated(spt04_hr_ss_agg_tbl)
```

### SPT05 aggregated plotting call

```{r}
plot_sample_sizes_aggregated(spt05_hr_ss_agg_tbl)
```

# Other

### Single Train/Test Split (commented out)

```{r}
spt04_full_tsibble <- as_tsibble(spt04_hr_daily_tbl, index=day)

# Set start and end of test window
train_size <- 25
spt04_train_start <- as.Date("2026-01-23")
spt04_test_start <- spt04_train_start + train_size

# Training on all data, restricted by phase
spt04_train_tsibble <- spt04_full_tsibble |> 
  filter(day >= spt04_train_start & day < spt04_test_start)


# Testing set
spt04_test_tsibble <- spt04_full_tsibble |> 
  filter(day >= spt04_test_start & day < spt04_test_start + 7) |>
  mutate(actual_hr_avg = hr_avg) |> select(-hr_avg)
```

```{r}
spt04_models <- fit_models(spt04_train_tsibble)

# Display order of Auto-ARIMA model
print(spt04_models["ARIMA"])

spt04_forecasts <- spt04_models |> forecast(h=7)
```

# Plotting Model Forecasts

### Benchmark Models

```{r}
# Convert to tibble for plotting and evaluation
spt04_mean_forecast_tbl <- spt04_forecasts |> filter(.model=='Mean') |> 
  as_tibble()

# Plot trend plus exponential smoothing curve plus forecast
spt04_forecasts |> filter(.model=='Mean') |>
  autoplot(filter(spt04_full_tsibble, 
                  day > spt04_train_start & day < spt04_test_start + 7)) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(spt04_models['Mean'])) +
  guides(colour = "none") + 
  labs(y="Daily Average Activity Count", x = "Day",
       title="Observed (Black) vs Mean Activity Counts")

# Close-up view of the week's predictions vs the actual data
ggplot() +
  geom_line(data = as_tibble(spt04_test_tsibble),
            aes(x = day, y = actual_hr_avg)) +
  geom_line(data = spt04_mean_forecast_tbl,
            aes(x = day, y = .mean), colour = "blue") +
  labs(y="Daily Average Activity Count", x = "Day", 
       title="Actual vs Mean Activity Counts for SPT05 June 02-08")
```

### Exponential Smoothing

```{r}
# Convert to tibble for plotting and evaluation
spt04_ets_forecast_tbl <- spt04_forecasts |> filter(.model=='ETS') |> 
  as_tibble()

# Plot trend plus exponential smoothing curve plus forecast
spt04_forecasts |> filter(.model=='ETS') |>
  autoplot(filter(spt04_full_tsibble, 
                  day > spt04_train_start & day < spt04_test_start + 7)) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(spt04_models['ETS'])) +
  guides(colour = "none") + 
  labs(y="Daily Average Activity Count", x = "Day",
       title="Observed (Black) vs ETS-Fitted (Orange) Activity Counts")

# Close-up view of the week's predictions vs the actual data
ggplot() +
  geom_line(data = as_tibble(spt04_test_tsibble),
            aes(x = day, y = actual_hr_avg)) +
  geom_line(data = spt04_ets_forecast_tbl,
            aes(x = day, y = .mean), colour = "blue") +
  labs(y="Daily Average Activity Count", x = "Day", 
       title="Actual vs ETS-Predicted Activity Counts for SPT05 June 02-08")
```

### ARIMA

```{r}
# Convert to tibble for plotting and evaluation
spt04_arima_forecast_tbl <- spt04_forecasts |> filter(.model=='ARIMA') |> 
  as_tibble()

# Plot trend plus exponential smoothing curve plus forecast
spt04_forecasts |> filter(.model=='ARIMA') |>
  autoplot(filter(spt04_full_tsibble, 
                  day > spt04_train_start & day < spt04_test_start + 7)) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(spt04_models['ARIMA'])) +
  guides(colour = "none")

# Close-up view of the week's predictions vs the actual data
ggplot() +
  geom_line(data = as_tibble(spt04_test_tsibble),
            aes(x = day, y = actual_hr_avg)) +
  geom_line(data = spt04_arima_forecast_tbl,
            aes(x = day, y = .mean), colour = "blue") +
  labs(y="Daily Average Activity Count", x = "Day", 
       title="Actual vs ARIMA-Predicted Activity Counts for SPT05 June 05-11")
```
